What We Eval in the Shadows

### 2021-04-13 analysis/environments.Rmd
Polymorphism in envir

nb_global_env_sites <- specific_env %>% filter(specific_env == "global") %>% .$n
nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n # Seems it is actually for envir == "new.env()"
if(length(nb_callee_env_sites) == 0) nb_callee_env_sites <- 0
nb_empty_env_sites <- specific_env %>% filter(specific_env == "empty") %>% .$n
if(length(nb_empty_env_sites) == 0) nb_empty_env_sites <- 0
nb_base_env_sites <- specific_env %>% filter(specific_env == "base") %>% .$n
if(length(nb_base_env_sites) == 0) nb_base_env_sites <- 0

@@

```{r}
nb_dataframe_env_sites <- nb_env_list_sites + nb_null_sites

nb_pack_sites <- nb_package_env_sites + nb_namespace_env_sites + nb_base_env_sites

@@

overview_table(
  r("nb dataframe env sites", nb_dataframe_env_sites),
  r("nb dataframe env site percent", ratio(nb_dataframe_env_sites, nb_call_sites)),
  r("nb package namespace env sites", nb_pack_sites),
  r("nb package namespace env site percent", ratio(nb_pack_sites, nb_call_sites)),
  r("nb function env sites", nb_function_env_sites),
  r("nb function env site percent", ratio(nb_function_env_sites, nb_call_sites))
)
```

>>>>

nb_global_env_sites <- specific_env %>% filter(specific_env == "global") %>% .$n
nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n # local, default envir argument
if(length(nb_callee_env_sites) == 0) nb_callee_env_sites <- 0
nb_empty_env_sites <- specific_env %>% filter(specific_env == "empty") %>% .$n
if(length(nb_empty_env_sites) == 0) nb_empty_env_sites <- 0
nb_base_env_sites <- specific_env %>% filter(specific_env == "base") %>% .$n
if(length(nb_base_env_sites) == 0) nb_base_env_sites <- 0

@@

```{r}

pack_envs <- env_classes %>% filter(is.na(new_env), is.na(hierarchy), specific_env != "global")
nb_pack_sites <- n_distinct(pack_envs$src_ref)

@@

synthetic_envs <- env_classes %>% filter(envir_type %in% c("list", "NULL") | !is.na(new_env) | specific_env == "empty")
nb_synthetic_sites <- n_distinct(synthetic_envs$src_ref)

global_envs <- env_classes %>% filter(is.na(new_env), specific_env == "global")
nb_strict_global_env_sites <- n_distinct(global_envs$src_ref)

overview_table(
  r("nb synthetic env sites", nb_synthetic_sites),
  r("nb synthetic env site percent", ratio(nb_synthetic_sites, nb_call_sites)),
  r("nb package namespace env sites", nb_pack_sites),
  r("nb package namespace env site percent", ratio(nb_pack_sites, nb_call_sites)),
  r("nb function env sites", nb_function_env_sites),
  r("nb function env site percent", ratio(nb_function_env_sites, nb_call_sites)),
  r("nb strict global env sites", nb_strict_global_env_sites),
  r("nb strict global env site percent", ratio(nb_strict_global_env_sites, nb_call_sites))
)
```

Polymorphism? How many sites have envir in several categories?

```{r}
pack_env_sites <- pack_envs %>% select(src_ref) %>% mutate(category = "package") 
function_env_sites <- bind_rows(select(caller_envs, src_ref), select(eval_calls_integer, src_ref)) %>% mutate(category = "function")
synthetic_env_sites <- synthetic_envs %>% select(src_ref) %>% mutate(category = "synthetic")
global_env_sites <- global_envs %>% select(src_ref)%>% mutate(category = "global")

category_per_site <- bind_rows(pack_env_sites, function_env_sites, synthetic_env_sites, global_env_sites) %>%
  group_by(src_ref) %>% summarise(n_categories = n_distinct(category))

poly_categories <- category_per_site %>% count(n_categories, sort = TRUE) %>%
   mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc))

one_category <- poly_categories %>% filter(n_categories == 1)

overview_table(
  r("nb one category envir sites", one_category$n),
  r("nb one category envir site percent", one_category$pc)
)

### 2021-04-13 paper/main.tex
Polymorphism in envir

\begin{itemize}
\item \emph{Function:} these environments correspond to some function active on
  the call stack and are obtained by functions such as \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} these are environments that built from data structures
  such as lists, dataframes, etc., or constructed explicitly with \c{new.env},
  \c{list2env} or \c{as.environment}.
\item \emph{Global:} the environment in which scripts or interactive commands
  are evaluated.
\item \emph{Package:} the environments of a loaded library.
\end{itemize}

As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a
function environment, with \c{global} as a distant second. This means that most
variable lookups and most side effects will take place amongst the local
variables of some function. But which function? Table~3 shows the offset from
the function that directly called \eval, thus 0 is the direct caller and 1 is
its parent. The data suggests that over 80\% of function environments are the
direct caller -- this means, from a user's point of view -- the environment of
the function where \eval occurs textually. It is interesting that some 1.5\% of
sites evaluate a code in an environment that is three frame or above from the
caller of \eval. This distance makes it more tricky to reason about the behavior
of code. As the actions of \eval may happen at a distance.

\begin{table}[h]\centering\small
\begin{minipage}{4cm}
  \begin{tabular}{r|r|r}\hline
 Environment & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
 Synthetic & ? & ? \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Environments per site} \label{tab:highlevelenvironments}
\end{minipage} \hspace{1cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{r|r|r}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{fn}
\end{minipage}\hspace{1cm}
\begin{minipage}{3.8cm}
\begin{tabular}{r|r|r} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} \label{tab:newenvs}
\caption{Wrapper envs.}
\end{minipage}
\end{table}

@@

Synthetic environments must be given a parent. The parent is used to lookup
variables that are not found the new environment (side effects go to the new
environment and not to its parent). Table 5 shows the parents of synthetic
environments. Over half are functions, the second most frequent case is packages.
??? Say more ????  FIX: so that we get to 100\%

@@

\begin{itemize}
\item {\it Promise:} Function arguments are passed as promises, suspended
  computations, from a promise one can retrieve the source expression with a
  call to the \c{substitute} function.
\item {\it Reflection:} The \c{match.call} function reflectively captures
  the expression that called the current function.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote} and \c{expression} functions.
\item {\it String:} Finally, expressions can be created from strings by invoking
  one of \c{parse}, \c{str2expression} and \c{str2lang}.
\end{itemize}

>>>>

\begin{compactitem}[---]
\item \emph{Function:} these environments correspond to some function active on
  the call stack and are obtained by functions such as \c{parent.frame()} or
  \c{sys.frame()}.
\item \emph{Synthetic:} these are environments that built from data structures
  such as lists, dataframes, etc., or constructed explicitly with \c{new.env},
  \c{list2env} or \c{as.environment}. It also includes the empty environment.
\item \emph{Global:} the environment in which scripts or interactive commands
  are evaluated.
\item \emph{Package:} the environments of a loaded library.
\end{compactitem}

As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a
function environment, with \c{global} as a distant second. This means that most
variable lookups and most side effects will take place amongst the local
variables of some function. But which function? Table~\ref{tab:funoffset} shows the offset from
the function that directly called \eval, thus 0 is the direct caller and 1 is
its parent. The data suggests that over 80\% of function environments are the
direct caller -- this means, from a user's point of view -- the environment of
the function where \eval occurs textually. It is interesting that some 1.5\% of
sites evaluate a code in an environment that is three frame or above from the
caller of \eval. This distance makes it more tricky to reason about the behavior
of code, as the actions of \eval may happen at a distance.

\begin{table}[h]\centering\small
\begin{minipage}{4cm}
  \begin{tabular}{r|r|r}\hline
 Environment & \#sites & \%sites \\\hline
 Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
 Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
 Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
 Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
\end{tabular}
\caption{Environments per site} \label{tab:highlevelenvironments}
\end{minipage} \hspace{1cm}
\begin{minipage}{3.7cm}\centering
\begin{tabular}{r|r|r}\hline
 Offset & \#sites & \%sites \\\hline
  \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
  \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
  \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
 \end{tabular}
\caption{Function offset}\label{tab:funoffset}
\end{minipage}\hspace{1cm}
\begin{minipage}{3.8cm}
\begin{tabular}{r|r|r} \hline
Parent & \#sites & \%sites \\\hline
Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
\end{tabular} \label{tab:newenvs}
\caption{Wrapper envs.}
\end{minipage}
\end{table}
One site could appear in several categories. However, most sites, \ie \packageNbOneCategoryEnvirSitePercent, have only kind of the above environments passed to them.

@@

Synthetic environments must be given a parent. When the \c{envir} argument is a list or a data frame, \eval also builds a new environment using the \c{enclos} argument as the parent. For \c{new.env}, the parent is passed explicitly at construction time. The parent is used to lookup
variables that are not found in the new environment (side effects go to the new
environment and not to its parent). Table~\ref{tab:newenvs} shows the parents of synthetic
environments. Over half are functions, the second most frequent case is packages.
??? Say more ????  FIX: so that we get to 100\%
\todo{It is over 100\% because one site can be passed different kinds of environments so it is a percentage over the number of sites in the dataset. I could also normalize to 100\% if it is more intuitive. In that case, the percentage would be over the sum of the rows (i.e. with duplicated sites).}

@@

\begin{compactitem}[---]
\item {\it Promise:} Function arguments are passed as promises, suspended
  computations, from a promise one can retrieve the source expression with a
  call to the \c{substitute} function.
\item {\it Reflection:} The \c{match.call} function reflectively captures
  the expression that called the current function.
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote} and \c{expression} functions.
\item {\it String:} Finally, expressions can be created from strings by invoking
  one of \c{parse}, \c{str2expression} and \c{str2lang}.
\end{compactitem}
